---
title: "Breast cancer survival prediction"
author: "Tong Su, Minghe Wang, Kai Tan, Yifei Chen"
date: "2024-12-19"
output: github_document
---

```{r setup, include=FALSE}
# Load required libraries
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(caret)
library(ROCR)
library(tidyr)
library(reshape2)
library(corrplot)
library(car)
library(glmnet)
library(mgcv)
library(boot)
library(ROSE)
```

```{r data_cleaning}
# Load dataset
data <- read.csv("./data/Project_2_data.csv")

# Clean dataset
data_cleaned <- data %>%
  mutate(
    Race = factor(Race),
    Marital.Status = factor(trimws(Marital.Status)),
    T.Stage = factor(T.Stage),
    N.Stage = factor(N.Stage),
    X6th.Stage = factor(X6th.Stage),
    differentiate = factor(differentiate),
    Grade = factor(trimws(Grade)),  # Trim spaces before factoring
    A.Stage = factor(A.Stage),
    Estrogen.Status = factor(Estrogen.Status),
    Progesterone.Status = factor(Progesterone.Status),
    Status = ifelse(Status == "Dead", 1, 0)
  ) %>% 
  mutate(
    T.Stage = as.numeric(factor(T.Stage, levels = c("T1", "T2", "T3", "T4"), ordered = TRUE)),
    N.Stage = as.numeric(factor(N.Stage, levels = c("N1", "N2", "N3"), ordered = TRUE)),
    X6th.Stage = as.numeric(factor(X6th.Stage, levels = c("IIA", "IIB", "IIIA", "IIIB", "IIIC"), ordered = TRUE)),
    Grade = as.numeric(factor(Grade, levels = c("1", "2", "3", "anaplastic; Grade IV"), ordered = TRUE)),
    A.Stage = as.numeric(factor(A.Stage, levels = c("Regional", "Distant"), ordered = TRUE))
  ) %>% 
  select(-differentiate)
```

`Grade` and `differentiate` are redundant variables, so we drop `differentiate`.

# 1. Descriptive Summary

```{r}
# summary_stats for numeric predictors
summary_numeric <- data_cleaned |> 
  select_if(is.numeric) |>
  summarise_all(list(
    mean = ~mean(.x, na.rm = TRUE), 
    median = ~median(.x, na.rm = TRUE), 
    sd = ~sd(.x, na.rm = TRUE),
    min = ~min(.x, na.rm = TRUE),
    max = ~max(.x, na.rm = TRUE),
    iqr = ~IQR(.x, na.rm = TRUE),
    count = ~sum(!is.na(.x))
  ))

summary_numeric_long <- summary_numeric |> 
  pivot_longer(
    cols = everything(),        
    names_to = c("Variable", "Statistic"), 
    names_sep = "_",            
    values_to = "Value"         
  )
summary_numeric_wide <- summary_numeric_long |> 
  pivot_wider(
    names_from = Statistic,    
    values_from = Value         
  )
summary_numeric_wide

# Missing data
missing_percentage <- data_cleaned |> 
  summarise(across(everything(), ~ mean(is.na(.)) * 100, .names = "missing_pct_{.col}"))
missing_percentage
```

# 2. Data Visualization

```{r}
# Distribution of continuous variables
numeric_vars <- data_cleaned |> select_if(is.numeric)

# Histogram for continuous variables
for (col in names(numeric_vars)) {
  ggplot(data_cleaned, aes_string(x = col)) +
    geom_histogram(bins = 30, fill = "blue", color = "black") +
    ggtitle(paste("Distribution of", col)) +
    theme_minimal()
}

# Scatter plot for continuous variables
predictors = c("Age", "Tumor.Size", "Regional.Node.Examined", "Reginol.Node.Positive")
sorted_data = melt(data_cleaned, id.vars = "Survival.Months", measure.vars = predictors)
ggplot(sorted_data, aes(x = value, y = Survival.Months)) +
  geom_point(alpha = 1) +
  geom_smooth(method = "lm", color = "blue") +
  facet_wrap(~variable, scales = "free_x") +  
  labs(title = "Scatter Plots of Predictors vs Survival Time",
       x = "Predictor Value",
       y = "Survival Time (Months)") +
  theme_minimal()

# Correlation heatmap with correlation coefficient
cor_matrix <- cor(numeric_vars, use = "complete.obs")
corrplot(cor_matrix, method = "number", type = "full", tl.col = "black", tl.srt = 45, title = "Correlation Heatmap")

# Bar Plot for categorical variables
categorical_vars <- names(data_cleaned)[sapply(data_cleaned, is.factor) | sapply(data_cleaned, is.character)]
if (length(categorical_vars) > 0) {
  for (var in categorical_vars) {
    p <- ggplot(data_cleaned, aes_string(x = var, fill = var)) +
      geom_bar() +
      labs(
        title = paste("Distribution of", var),
        x = var,
        y = "Count"
      ) +
      theme_minimal() +
      scale_fill_brewer(palette = "Set2")  
    
    print(p)  
  }
} else {
  message("No categorical variables found in the dataset.")
}

#show pairs with strong correlation
for (i in 1:(nrow(cor_matrix) - 1)) {
  for (j in (i + 1):ncol(cor_matrix)) {
    if (abs(cor_matrix[i, j]) > 0.7) {
      cat("Pairs with strong correlation:", rownames(cor_matrix)[i], "~", colnames(cor_matrix)[j], 
          "with correlation =", cor_matrix[i, j], "\n")
    }
  }
}

#show pairs with weak correlation
for (i in 1:(nrow(cor_matrix) - 1)) {
  for (j in (i + 1):ncol(cor_matrix)) {
    if (abs(cor_matrix[i, j]) > 0.2 & abs(cor_matrix[i, j]) < 0.4) {
      cat("Pairs with weak correlation:", rownames(cor_matrix)[i], "~", colnames(cor_matrix)[j], 
          "with correlation =", cor_matrix[i, j], "\n")
    }
  }
}
```

# 3. Model Building: Survival Analysis

```{r}
# Build Cox Proportional Hazards model
cox_model <- coxph(Surv(Survival.Months, Status) ~ ., data = data_cleaned)
summary(cox_model)

# Save model summary
tidy_cox <- broom::tidy(cox_model)
write.csv(tidy_cox, "cox_model_summary.csv")

# Evaluate Model Performance
# Concordance index
c_index <- summary(cox_model)$concordance
cat("Concordance Index:", c_index, "\n")
```

## Model Diagnositics

```{r}
cox_zph <- cox.zph(cox_model)
print(cox_zph)

# Fit Cox model with stratification on violating covariates
cox_model_strat <- coxph(Surv(Survival.Months, Status) ~ Age + Race + Marital.Status + T.Stage + 
                          N.Stage + X6th.Stage + Grade + Tumor.Size + 
                          Regional.Node.Examined + Reginol.Node.Positive + 
                          strata(A.Stage) + strata(Estrogen.Status) + 
                          strata(Progesterone.Status), 
                        data = data_cleaned)

summary(cox_model_strat)

cox_zph_strat <- cox.zph(cox_model_strat)
print(cox_zph_strat)

```

## Model Selection

```{r coxPH_model_selection_stepwise}
full_model <- coxph(Surv(Survival.Months, Status) ~ Age + Race + Marital.Status + T.Stage + 
                     N.Stage + X6th.Stage + Grade + Tumor.Size + 
                     Regional.Node.Examined + Reginol.Node.Positive + 
                     strata(A.Stage) + strata(Estrogen.Status) + 
                     strata(Progesterone.Status), 
                   data = data_cleaned)

# Define the null model with only stratified variables
null_model <- coxph(Surv(Survival.Months, Status) ~ strata(A.Stage) + strata(Estrogen.Status) + 
                     strata(Progesterone.Status), 
                   data = data_cleaned)

step_backward <- MASS::stepAIC(full_model, 
                          scope = list(lower = null_model, upper = full_model), 
                          direction = "backward")
summary(step_backward)
```

Here we use step forward selection with AIC to obtain an insight on effective predictor, and `Marital.Status` and `X6th.Stage` are removed according to the summary. Then we perform LASSO regularization to find most influential predictors in our Cox PH model.

```{r coxPH_model_selection_lasso}
# Define stratified variables
strata_vars <- c("A.Stage", "Estrogen.Status", "Progesterone.Status")

# Create predictor matrix excluding stratified variables
x <- model.matrix(~ Age + Race + T.Stage + 
                   N.Stage + Grade + Tumor.Size + 
                   Regional.Node.Examined + Reginol.Node.Positive, 
                 data = data_cleaned)[, -1]  # Remove intercept

# Define the response variable
y <- Surv(time = data_cleaned$Survival.Months, event = data_cleaned$Status)

# Set seed for reproducibility
set.seed(123)

# Fit LASSO Cox model with cross-validation
lasso_cox <- cv.glmnet(x, y, family = "cox", alpha = 1, 
                       nfolds = 10, 
                       standardize = TRUE)

# Plot cross-validated partial likelihood deviance
plot(lasso_cox)
title("LASSO Cox Model Cross-Validation", line = 2.5)

# Optimal lambda values
optimal_lambda_min <- lasso_cox$lambda.min

# Extract coefficients at lambda.min
coef_min <- coef(lasso_cox, s = "lambda.min")
selected_vars_min <- rownames(coef_min)[which(coef_min != 0)]
selected_vars_min <- selected_vars_min[selected_vars_min != "(Intercept)"]  # Remove intercept

print("Selected Variables at lambda.min:")
print(selected_vars_min)

# Fit the final Cox model
final_cox_lasso <- coxph(Surv(Survival.Months, Status) ~ Age + Race + T.Stage + N.Stage
                         + Grade + Regional.Node.Examined + Reginol.Node.Positive + strata(A.Stage) + 
                           strata(Estrogen.Status) + strata(Progesterone.Status), data = data_cleaned)
summary(final_cox_lasso)
```


```{r cv_for_coxPH_final, warning = FALSE}
library(timeROC)
library(rms)
# Combine stratified variables into a single strata identifier
data_cleaned <- data_cleaned %>%
  mutate(Combined_Strata = paste(A.Stage, Estrogen.Status, Progesterone.Status, sep = "_"))

# Set seed for reproducibility
set.seed(123)

# Create 5 stratified folds
folds <- createFolds(data_cleaned$Combined_Strata, k = 5, list = TRUE, returnTrain = FALSE)

# Initialize vector to store C-Index for each fold
c_indices <- c()

# Loop through each fold
for(i in 1:length(folds)) {
  # Define training and validation indices
  val_indices <- folds[[i]]
  train_indices <- setdiff(1:nrow(data_cleaned), val_indices)
  
  # Split the data
  train_data <- data_cleaned[train_indices, ]
  val_data <- data_cleaned[val_indices, ]
  
  # Fit the Cox model on training data
  cox_model_cv <- coxph(Surv(Survival.Months, Status) ~ Age  + Race + T.Stage + N.Stage
                        + Grade + Reginol.Node.Positive + Regional.Node.Examined +
                          strata(A.Stage) + strata(Estrogen.Status) + 
                          strata(Progesterone.Status), 
                        data = train_data)
  
  # Predict risk scores on validation data
  risk_scores <- predict(cox_model_cv, newdata = val_data, type = "risk")
  
  # Calculate C-Index for validation set
  concordance_val <- survConcordance(Surv(val_data$Survival.Months, val_data$Status) ~ risk_scores)
  c_indices <- c(c_indices, concordance_val$concordance)
}

# Calculate average C-Index across folds
average_c_index <- mean(c_indices)
print(paste("Average C-Index across 5 folds:", round(average_c_index, 3)))

```


# 4. Model Building: Logistic Regression

## Varibales selection

```{r}
# Building model
logistic_full <- glm(Status ~ . + T.Stage*Tumor.Size + N.Stage*Reginol.Node.Positive, family = binomial, data = data_cleaned)
logistic_null <- glm(Status ~ 1, family = binomial, data = data_cleaned)

# Bootstrap for Logistic Full Model
boot_fn <- function(data, indices) {
  data_boot <- data[indices, ]
  model <- glm(Status ~ ., family = binomial, data = data_boot)
  return(coef(model))
}

set.seed(123)
boot_results <- boot(data_cleaned, boot_fn, R = 1000)
print(boot_results)

# backward selection
backward_model <- step(logistic_full, direction = "backward")

# forward selection
forward_model <- step(logistic_null, scope = list(lower = logistic_full, upper = logistic_null), direction = "forward")

# stepwise selection
stepwise <- step(logistic_null, scope = list(lower = logistic_full, upper = logistic_null), direction = "both")

# examine if the same model was build
formula(backward_model)
formula(forward_model)
formula(stepwise)

# check if there is a close call
summary(backward_model)
summary(forward_model)
summary(stepwise)

# results
results <- data.frame(
Model = c("forward", "backward", "stepwise"),
AIC = c(AIC(forward_model), AIC(backward_model), AIC(stepwise)), 
BIC = c(BIC(forward_model), BIC(backward_model), BIC(stepwise))
) 
results

#compare AIC with and without Estrogen.StatusPosive
Estrogen.Status_included_backward = glm(Status ~ Age + Race + T.Stage + N.Stage + Grade + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Reginol.Node.Positive + Survival.Months, family = binomial, data = data_cleaned)

Estrogen.Status_excluded_backward = glm(Status ~ Age + Race + T.Stage + N.Stage + Grade + Progesterone.Status + Regional.Node.Examined + Reginol.Node.Positive + Survival.Months, family = binomial, data = data_cleaned) 

result1 <- data.frame(
Model = c("with Estrogen.StatusPosive", "without Estrogen.StatusPosive"),
AIC = c(AIC(Estrogen.Status_included_backward), AIC(Estrogen.Status_excluded_backward)), 
BIC = c(BIC(Estrogen.Status_included_backward), BIC(Estrogen.Status_excluded_backward))
) 
result1

#compare AIC with and without N.Stage*Regional.Node.Positive
Interaction_included_backward = glm(Status ~ Age + Race + T.Stage + N.Stage + Grade + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Reginol.Node.Positive + Survival.Months + N.Stage*Reginol.Node.Positive, family = binomial, data = data_cleaned)

Interaction_excluded_backward = glm(Status ~ Age + Race + T.Stage + N.Stage + Grade + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Reginol.Node.Positive + Survival.Months, family = binomial, data = data_cleaned)

result2 <- data.frame(
Model = c("with N.Stage*Regional.Node.Positive", "without N.Stage*Regional.Node.Positive"),
AIC = c(AIC(Interaction_included_backward), AIC(Interaction_excluded_backward)), 
BIC = c(BIC(Interaction_included_backward), BIC(Interaction_excluded_backward))
) 
result2

#compare AIC with and without N.Stage
Nstage_included_backward = glm(Status ~ Age + Race + T.Stage + N.Stage + Grade + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Reginol.Node.Positive + Survival.Months, family = binomial, data = data_cleaned)

Nstage_excluded_backward = glm(Status ~ Age + Race + T.Stage + Grade + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Reginol.Node.Positive + Survival.Months, family = binomial, data = data_cleaned)

result3 <- data.frame(
Model = c("with N.Stage", "without N.Stage"),
AIC = c(AIC(Nstage_included_backward), AIC(Nstage_excluded_backward)), 
BIC = c(BIC(Nstage_included_backward), BIC(Nstage_excluded_backward))
) 
result3

#compare AIC with and without Regional.Node.Positive
RNP_included_backward = glm(Status ~ Age + Race + T.Stage + N.Stage + Grade + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Reginol.Node.Positive + Survival.Months, family = binomial, data = data_cleaned)

RNP_excluded_backward = glm(Status ~ Age + Race + T.Stage + N.Stage + Grade + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Survival.Months, family = binomial, data = data_cleaned)

result4 <- data.frame(
Model = c("with Regional.Node.Positive", "without Regional.Node.Positive"),
AIC = c(AIC(RNP_included_backward), AIC(RNP_excluded_backward)), 
BIC = c(BIC(RNP_included_backward), BIC(RNP_excluded_backward))
) 
result4
```

## try using lasso for model selection and test the model 

```{r}
predictors <- model.matrix(Status ~ ., data = data_cleaned)[, -1]
patients_status <- as.numeric(data_cleaned$Status)
lasso_model <- cv.glmnet(
  predictors, 
  patients_status, 
  alpha = 1,            
  family = "binomial"   
)

# best lambda
best_lambda <- lasso_model$lambda.min
best_lambda

# select variables
lasso_coefs <- coef(lasso_model, s = best_lambda)
lasso_coefs_matrix <- as.matrix(lasso_coefs)
selected_dummies <- rownames(lasso_coefs_matrix)[lasso_coefs_matrix != 0][-1] 

selected_predictors <- unique(gsub("(.*)\\..*", "\\1", selected_dummies))
selected_predictors

#logistics regression
logistics_test2 <- glm(Status ~ Age + Race + Marital.Status + T.Stage + N.Stage + Grade + Tumor.Size + Estrogen.Status + Progesterone.Status + Regional.Node.Examined + Survival.Months, data = data_cleaned, family = binomial)
summary(logistics_test2)

#diagosis
std_residuals <- rstandard(logistics_test2)
leverage <- hatvalues(logistics_test2)
cooks_distance <- cooks.distance(logistics_test2)
plot_data <- data.frame(Leverage = leverage, Residuals = std_residuals)

ggplot(plot_data, aes(x = leverage, y = std_residuals)) +
  geom_point(aes(color = cooks_distance > 4 / nrow(data_cleaned))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 2 * mean(leverage), linetype = "dashed", color = "blue") +
  labs(
    title = "Residuals vs Leverage with Cook's Distance",
    x = "Leverage (Hat Values)",
    y = "Standardized Residuals (Deviance)"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("black", "purple"), labels = c("Not Influential", "Influential"))
```

## Likelihood Ratio Test

```{r}
#Regional.Node.Examined * Reginol.Node.Positive (significant, lrt suggests they should be included)
interaction_model <- glm(Status ~ Regional.Node.Examined * Reginol.Node.Positive, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ Regional.Node.Examined + Reginol.Node.Positive, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)

#A.Stage * Reginol.Node.Positive (significant, lrt suggests they should be included)
interaction_model <- glm(Status ~ A.Stage * Reginol.Node.Positive, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ A.Stage + Reginol.Node.Positive, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)

#T.Stage * Tumor.Size
interaction_model <- glm(Status ~ T.Stage * Tumor.Size, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ T.Stage + Tumor.Size, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)

#N.Stage * X6th.Stage
interaction_model <- glm(Status ~ N.Stage*X6th.Stage, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ N.Stage + X6th.Stage, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)

#N.Stage * Reginol.Node.Positive
interaction_model <- glm(Status ~ N.Stage*Reginol.Node.Positive, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ N.Stage + Reginol.Node.Positive, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)

#X6th.Stage ~ Reginol.Node.Positive
interaction_model <- glm(Status ~ X6th.Stage*Reginol.Node.Positive, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ X6th.Stage + Reginol.Node.Positive, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)

#T.Stage ~ X6th.Stage
interaction_model <- glm(Status ~ T.Stage*X6th.Stage, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ T.Stage + X6th.Stage, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)


#T.Stage ~ N.Stage
interaction_model <- glm(Status ~ T.Stage * N.Stage, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ T.Stage + N.Stage, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)

#T.Stage ~ A.Stage
interaction_model <- glm(Status ~ T.Stage * A.Stage, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ T.Stage + A.Stage, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)

#A.Stage * Reginol.Node.Positive !!!
interaction_model <- glm(Status ~ A.Stage * Reginol.Node.Positive, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ A.Stage + Reginol.Node.Positive, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)

#Tumor.Size ~ Reginol.Node.Positive
interaction_model <- glm(Status ~ X6th.Stage * Regional.Node.Examined, data = data_cleaned, family = binomial)
base_model <- glm(Status ~ X6th.Stage + Regional.Node.Examined, data = data_cleaned, family = binomial)
lrt <- anova(base_model, interaction_model, test = "LRT")
print(lrt)
```

## logistic regression

```{r}
# with Reginol.Node.Positive and Regional.Node.Examined * Reginol.Node.Positive
logistics <- glm(Status ~ Age + Race + T.Stage + N.Stage + Grade + 
    Estrogen.Status + Progesterone.Status + Regional.Node.Examined + 
    Reginol.Node.Positive + Regional.Node.Examined * 
    Reginol.Node.Positive, family = binomial, data = data_cleaned)
summary(logistics)

predicted_probabilities <- predict(logistics, newdata = data_cleaned, type = "response")
data$predicted_status <- ifelse(predicted_probabilities > 0.5, "Alive", "Dead")
```

# cv for glm
```{r}
#cv for glm
data_cleaned$Status <- factor(data_cleaned$Status, levels = c(0, 1))
control <- trainControl(method = "cv", number = 10)
logistic_cv_model <- train(
  Status ~Age + Race + T.Stage + N.Stage + Grade + 
    Estrogen.Status + Progesterone.Status + Regional.Node.Examined + 
    Reginol.Node.Positive + Regional.Node.Examined * 
    Reginol.Node.Positive,
  data = data_cleaned,
  method = "glm",
  family = binomial,
  trControl = control
)
logistic_cv_model


```

#GAM
```{r}

#GAM using predictors from auto procedure
logistics_gam_auto <- gam(Status ~ Age + Race + T.Stage + N.Stage + Grade + 
    Estrogen.Status + Progesterone.Status + Regional.Node.Examined + 
    Reginol.Node.Positive + Regional.Node.Examined * 
    Reginol.Node.Positive,
    data = data_cleaned,
    family = binomial)
summary(logistics_gam_auto)
AIC(logistics_gam_auto, logistics)

#Gam using predictors from LASSO
logistics_gam_lasso <- gam(Status ~ s(Age) + Race + Marital.Status + T.Stage + N.Stage + Grade + s(Tumor.Size) + Estrogen.Status + 
    Progesterone.Status + s(Regional.Node.Examined), data = data_cleaned, family = binomial)

summary(logistics_gam_lasso)
AIC(logistics_gam_auto, logistics_test2)
   


```

#cv for GAM
```{r}

```


## Model Diagnostics

```{r logit_diagnostics}
# Test for linearity
test_logit_model <- glm(Status ~ ., family = binomial, data = data_cleaned)

probabilities <- predict(test_logit_model, type = "response")
mydata <- data_cleaned %>%
  dplyr::select(Age, Tumor.Size, Regional.Node.Examined, Reginol.Node.Positive)

predictors <- colnames(mydata)
# Bind the logit and tidying the data for plot
mydata <- mydata %>%
  mutate(logit = log(probabilities/(1-probabilities))) %>%
  gather(key = "predictors", value = "predictor.value", -logit)

ggplot(mydata, aes(logit, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_wrap(~predictors, scales = "free_y")

logit_vif_values <- vif(test_logit_model)
print(logit_vif_values)
data_cleaned <- data_cleaned %>% 
  select(-X6th.Stage)
test_logit_model2 <- glm(Status ~ ., family = binomial, data = data_cleaned)
logit_vif_values2 <- vif(test_logit_model2)
print(logit_vif_values2)
```

```{r}
# independence of errors
residuals <- residuals(test_logit_model2, type = "pearson")
plot(residuals, main = "Residual Plot", xlab = "Index", ylab = "Residuals")
abline(h = 0, col = "red")

#outlier assumption
plot(test_logit_model2, which = 4, id.n = 3)
model.data <- broom::augment(test_logit_model2) %>% 
  mutate(index = 1:n()) 
model.data %>% top_n(3, .cooksd)
ggplot(model.data, aes(index, .std.resid)) + 
  geom_point(aes(color = Status), alpha = .5) +
  theme_bw()
model.data %>% 
  filter(abs(.std.resid) > 3)

# existence of of strongly influential outliers
std_residuals <- rstandard(test_logit_model2)
leverage <- hatvalues(test_logit_model2)
cooks_distance <- cooks.distance(test_logit_model2)
plot_data <- data.frame(Leverage = leverage, Residuals = std_residuals)

ggplot(plot_data, aes(x = leverage, y = std_residuals)) +
  geom_point(aes(color = cooks_distance > 4 / nrow(data_cleaned))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 2 * mean(leverage), linetype = "dashed", color = "blue") +
  labs(
    title = "Residuals vs Leverage with Cook's Distance",
    x = "Leverage (Hat Values)",
    y = "Standardized Residuals (Deviance)"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("black", "purple"), labels = c("Not Influential", "Influential"))

```

Here we diagnose the logistric regression model by testing its assumption. According to the results, we found non-linear relationship between continuous variables and the log odd of binary outcome in our initial model. There is multicolinearity for `T.Stage`, `N.Stage`, `X6th.Stage`. However, we found that these variables might be highly correlated since they are using similar measuring systems. By removing `X6th.Stage`, the multicolinearity no longer exist. And out initial model pass the test for error independence and influencial outliers.

# 4. Fairness Analysis
# 4. Fairness Analysis
```{r}
# Define demographic groups to evaluate
group_var <- "Race"  # Example: Race
unique_groups <- unique(data_cleaned[[group_var]])
# Initialize results dataframe
fairness_results <- data.frame(
  Group = character(),
  Accuracy = numeric(),
  Precision = numeric(),
  Recall = numeric(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)
# Loop through each group and calculate metrics
for (group in unique_groups) {
  subset_data <- data_cleaned[data_cleaned[[group_var]] == group, ]
  
  # Make predictions
  probabilities <- predict(logistics_test2, newdata = subset_data, type = "response")
  predictions <- ifelse(probabilities > 0.5, 1, 0)
  # True labels
  true_labels <- subset_data$Status
  
predictions <- as.numeric(as.character(predictions))  # Convert from factor to numeric
true_labels <- as.numeric(as.character(true_labels))  # Convert from factor to numeric
accuracy <- mean(predictions == true_labels)
precision <- sum(predictions == 1 & true_labels == 1) / sum(predictions == 1)
recall <- sum(predictions == 1 & true_labels == 1) / sum(true_labels == 1)
# Add to results
  fairness_results <- rbind(
    fairness_results,
    data.frame(
      Group = group,
      Accuracy = accuracy,
      Precision = precision,
      Recall = recall
    )
  )
}
# Display results
print(fairness_results)
```
```{r}
# Calculate positive prediction rates for each group
positive_rates <- data_cleaned %>%
  mutate(
    Predicted = predict(logistics_test2, type = "response") > 0.5
  ) %>%
  group_by(Race) %>%
  summarise(
    PositiveRate = mean(Predicted)
  )
# Display positive rates
print(positive_rates)
ggplot(positive_rates, aes(x = Race, y = PositiveRate, fill = Race)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Positive Prediction Rates by Race",
    x = "Race",
    y = "Positive Prediction Rate"
  ) +
  theme_minimal()
```
```{r}
# Calculate recall (true positive rate) for each group
recall_rates <- data_cleaned %>%
  mutate(
    Predicted = predict(logistics_test2, type = "response") > 0.5
  ) %>%
  group_by(Race) %>%
  summarise(
    Recall = sum(Predicted & Status == 1) / sum(Status == 1)
  )
# Display recall rates
print(recall_rates)
```
```{r}
# Calculate precision for each group
precision_rates <- data_cleaned %>%
  mutate(
    Predicted = predict(logistics_test2, type = "response") > 0.5
  ) %>%
  group_by(Race) %>%
  summarise(
    Precision = sum(Predicted & Status == 1) / sum(Predicted)
  )
# Display precision rates
print(precision_rates)
```
```{r}
# Combine "Black" and "Other" into one category
data_cleaned <- data_cleaned %>%
  mutate(
    Race_Grouped = ifelse(Race %in% c("Black", "Other"), "Black/Other", "White")
  )
# Check the new distribution of Race_Grouped
table(data_cleaned$Race_Grouped)
```
```{r}
# Define the grouped Race variable
group_var <- "Race_Grouped"
unique_groups <- unique(data_cleaned[[group_var]])
# Initialize results dataframe
fairness_results_grouped <- data.frame(
  Group = character(),
  Accuracy = numeric(),
  Precision = numeric(),
  Recall = numeric(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)
# Loop through each group and calculate metrics
for (group in unique_groups) {
  subset_data <- data_cleaned[data_cleaned[[group_var]] == group, ]
  # Make predictions
  probabilities <- predict(logistics_test2, newdata = subset_data, type = "response")
  predictions <- ifelse(probabilities > 0.5, 1, 0)
  # True labels
  true_labels <- subset_data$Status
  
predictions <- as.numeric(as.character(predictions))  # Convert from factor to numeric
true_labels <- as.numeric(as.character(true_labels))  # Convert from factor to numeric
accuracy <- mean(predictions == true_labels)
precision <- sum(predictions == 1 & true_labels == 1) / sum(predictions == 1)
recall <- sum(predictions == 1 & true_labels == 1) / sum(true_labels == 1)
  
  # Add to results
  fairness_results_grouped <- rbind(
    fairness_results_grouped,
    data.frame(
      Group = group,
      Accuracy = accuracy,
      Precision = precision,
      Recall = recall
    )
  )
}
# Display results
print(fairness_results_grouped)
```
```{r}
# Calculate positive prediction rates for each group
positive_rates_grouped <- data_cleaned %>%
  mutate(
    Predicted = predict(logistics_test2, type = "response") > 0.5
  ) %>%
  group_by(Race_Grouped) %>%
  summarise(
    PositiveRate = mean(Predicted)
  )
# Display positive rates
print(positive_rates_grouped)
```
```{r}
# Calculate recall (true positive rate) for each group
recall_rates_grouped <- data_cleaned %>%
  mutate(
    Predicted = predict(logistics_test2, type = "response") > 0.5
  ) %>%
  group_by(Race_Grouped) %>%
  summarise(
    Recall = sum(Predicted & Status == 1) / sum(Status == 1)
  )
# Display recall rates
print(recall_rates_grouped)
```
```{r}
# Calculate precision for each group
precision_rates_grouped <- data_cleaned %>%
  mutate(
    Predicted = predict(logistics_test2, type = "response") > 0.5
  ) %>%
  group_by(Race_Grouped) %>%
  summarise(
    Precision = sum(Predicted & Status == 1) / sum(Predicted)
  )
# Display precision rates
print(precision_rates_grouped)
```
```{r}
# Bar plot for positive prediction rates
ggplot(positive_rates_grouped, aes(x = Race_Grouped, y = PositiveRate, fill = Race_Grouped)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Positive Prediction Rates by Race Group",
    x = "Race Group",
    y = "Positive Prediction Rate"
  ) +
  theme_minimal()
```
```{r}
# Perform chi-squared test
chisq_test_grouped <- chisq.test(positive_rates_grouped$PositiveRate)
print(chisq_test_grouped)
```

## confusion matrix
```{r}
data_white <- data_cleaned %>% filter(Race == "White")
data_other <- data_cleaned %>% filter(Race %in% c("Other", "Black"))

data_white$predicted_probabilities <- predict(logistics, newdata = data_white, type = "response")
data_other$predicted_probabilities <- predict(logistics, newdata = data_other, type = "response")

data_white$predicted_status <- ifelse(data_white$predicted_probabilities > 0.5, 1, 0)
data_other$predicted_status <- ifelse(data_other$predicted_probabilities > 0.5, 1, 0)

confusion_white <- confusionMatrix(
  as.factor(data_white$predicted_status),
  as.factor(data_white$Status)
)
confusion_other <- confusionMatrix(
  as.factor(data_other$predicted_status),
  as.factor(data_other$Status)
)

confusion_white
confusion_other
```

